\documentclass[a4paper,11pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{GKS-I Formal Proof: A Mathematical Guide to the Lean 4 Code}
\author{ising-model project}
\date{2026-04-10}

\begin{document}
\maketitle

\begin{abstract}
This document provides a mathematical walkthrough of the Lean~4
formalization of the first Griffiths inequality (GKS-I) for
the finite Ising model. Each section corresponds to a Lean source file
and lists the relevant definitions and theorems. The purpose is to
make the formal proof readable as ordinary mathematics.

The proof follows Glimm--Jaffe~\cite{glimm-jaffe}, Theorem~4.1.1
(pp.~56--57), adapted to the discrete $\{+1,-1\}$ Ising model.
The ``non-negative correlations'' (HNC) framework and the cosh/sinh
decomposition are standard; see also
Friedli--Velenik~\cite{friedli-velenik}, Theorem~3.22 (pp.~99--100).
\end{abstract}

\tableofcontents

\section{Basic definitions (\texttt{Basic.lean})}

\texttt{Basic.lean} contains all fundamental types (Spin, Config,
IsingParams, Ferromagnetic) as well as spin operations (flip,
flipAt, sign, toSign) and their basic properties.

\begin{definition}[Spin]
$\mathrm{Spin} = \{\mathrm{up}, \mathrm{down}\}$ with sign function
$s : \mathrm{Spin} \to \mathbb{Z}$, $s(\mathrm{up})=+1$,
$s(\mathrm{down})=-1$.
\end{definition}

\noindent\textbf{Lean:} \texttt{inductive Spin}, \texttt{Spin.toSign}

\begin{definition}[Configuration]
A configuration on a type $\iota$ is a function
$\sigma : \iota \to \mathrm{Spin}$.
\end{definition}

\noindent\textbf{Lean:} \texttt{abbrev Config ($\iota$) := $\iota$ $\to$ Spin}

\begin{definition}[Ising parameters]
Over an ordered field $K$: coupling $J : K$, external field $h : K$,
inverse temperature $\beta : K$.
Ferromagnetic condition: $J \geq 0$, $h \geq 0$, $\beta > 0$.
\end{definition}

\noindent\textbf{Lean:} \texttt{structure IsingParams},
\texttt{structure Ferromagnetic}

\section{Hamiltonian (\texttt{Hamiltonian.lean})}

\begin{definition}[Ising Hamiltonian]
For a simple graph $G$ on $\iota$,
\[
  H_G(\sigma) = -J \sum_{\{i,j\} \in E(G)} s(\sigma_i)\,s(\sigma_j)
  - h \sum_{i \in \iota} s(\sigma_i).
\]
\end{definition}

\noindent\textbf{Lean:} \texttt{interactionEnergy} $+$
\texttt{externalFieldEnergy} $=$ \texttt{hamiltonian}

Key auxiliary definitions (in \texttt{Basic.lean}):
\begin{itemize}
  \item \texttt{Spin.flip}: $\mathrm{up} \leftrightarrow \mathrm{down}$
  \item \texttt{Spin.sign $K$ $s$}: cast $s(\cdot)$ into the field $K$
  \item \texttt{Config.flip}, \texttt{Config.flipAt}: global and single-site spin flip
\end{itemize}
Key auxiliary definitions (in \texttt{Hamiltonian.lean}):
\begin{itemize}
  \item \texttt{edgeSpin}: $s(\sigma_i) \cdot s(\sigma_j)$ for an edge
    $\{i,j\}$, defined via \texttt{Sym2.lift}
\end{itemize}

\begin{lemma}[Spin flip symmetry --- \texttt{hamiltonian\_flip\_eq}]
If $h = 0$, then $H_G(\bar{\sigma}) = H_G(\sigma)$ where
$\bar{\sigma}_i = \mathrm{flip}(\sigma_i)$.
\end{lemma}

\begin{proof}
Each edge factor satisfies $(-s_i)(-s_j) = s_i s_j$
(\texttt{edgeSpin\_flip}), and the field term vanishes since $h=0$.
\end{proof}

\section{Partition function and Gibbs measure (\texttt{GibbsMeasure.lean})}

\begin{definition}[Partition function]
$Z_G = \sum_{\sigma : \iota \to \mathrm{Spin}} \exp(-\beta\, H_G(\sigma))$.
\end{definition}

\begin{theorem}[$Z > 0$ --- \texttt{partitionFunction\_pos}]
$Z_G > 0$.
\end{theorem}

\begin{proof}
Each summand $\exp(\cdot) > 0$ (\texttt{Real.exp\_pos}), and the
index set $\iota \to \mathrm{Spin}$ is nonempty
(\texttt{Finset.univ\_nonempty}).
Apply \texttt{Finset.sum\_pos}.
\end{proof}

\begin{definition}[Gibbs expectation, correlation]
$\langle F \rangle = Z^{-1} \sum_\sigma F(\sigma)\,e^{-\beta H}$, \quad
$\langle \sigma^A \rangle = \langle \prod_{i \in A} s(\sigma_i) \rangle$.
\end{definition}

\noindent\textbf{Lean:} \texttt{gibbsExpectation},
\texttt{spinProduct}, \texttt{correlation}

\subsection{Spin product algebra}

\begin{lemma}[\texttt{spinProduct\_sq}]
$(\sigma^A)^2 = 1$ for any $A$.
\end{lemma}

\begin{lemma}[\texttt{spinProduct\_mul}]
$\sigma^A \cdot \sigma^C = \sigma^{A \bigtriangleup C}$.
\end{lemma}

\begin{proof}
Write $\sigma^A = \prod_{i \in \iota}
(\text{if } i \in A \text{ then } s_i \text{ else } 1)$
(using \texttt{Finset.prod\_filter}).
Multiply pointwise (\texttt{Finset.prod\_mul\_distrib}).
At each site: if $i \in A \cap C$, $s_i \cdot s_i = 1$ (by
\texttt{Spin.toSign\_sq}); otherwise the factor is $s_i$ or $1$,
matching $A \bigtriangleup C$.
\end{proof}

\section{GKS-I proof (\texttt{Inequalities/GKS.lean})}

\subsection{Reduction to non-negative correlations}

\begin{theorem}[GKS-I --- \texttt{gks\_first}]
For ferromagnetic parameters, $\langle \sigma^A \rangle \geq 0$
for all $A$.
\end{theorem}

The proof factors as:
\begin{enumerate}
  \item Since $Z > 0$, it suffices to show the numerator
    $\sum_\sigma \sigma^A e^{-\beta H} \geq 0$
    (\texttt{gibbsExpectation\_nonneg\_of\_numerator\_nonneg}).
  \item This follows from a property of $e^{-\beta H}$ called
    \emph{non-negative correlations}
    (\texttt{gks\_numerator\_nonneg}).
\end{enumerate}

\subsection{Non-negative correlations (HNC)}

\begin{definition}[\texttt{HasNonnegCorrelations}]
A function $f : (\iota \to \mathrm{Spin}) \to \mathbb{R}$ has
\textbf{non-negative correlations} (HNC) if
$\sum_\sigma \sigma^S \cdot f(\sigma) \geq 0$ for every
$S \subseteq \iota$.
\end{definition}

\begin{lemma}[Base case --- \texttt{hasNonnegCorrelations\_one}]
$f \equiv 1$ has HNC.
\end{lemma}

\begin{proof}
For $S = \emptyset$: $\sum_\sigma 1 = 2^{|\iota|} > 0$.
For $S \neq \emptyset$: $\sum_\sigma \sigma^S = 0$.

The vanishing is proved via the involution
$\mathrm{flipAt}(j)$ for any $j \in S$
(\texttt{sum\_config\_spinProduct\_eq\_zero}).
Each pair $(\sigma, \mathrm{flipAt}(j, \sigma))$ contributes
$\sigma^S + (-\sigma^S) = 0$ because flipping at $j \in S$ negates
the spin product (\texttt{spinProduct\_flipAt\_neg}).
The involution is applied via \texttt{Finset.sum\_ninvolution}.
\end{proof}

\begin{lemma}[Preservation --- \texttt{hasNonnegCorrelations\_mul}]
\label{lem:mul}
If $f$ has HNC and $a, b \geq 0$, then
$\sigma \mapsto f(\sigma)(a + b\,\sigma^C)$ has HNC.
\end{lemma}

\begin{proof}
\[
  \sum_\sigma \sigma^S f(\sigma)(a + b\,\sigma^C)
  = a \underbrace{\sum_\sigma \sigma^S f(\sigma)}_{\geq 0}
  + b \underbrace{\sum_\sigma \sigma^{S \bigtriangleup C} f(\sigma)}_{\geq 0}
\]
using $\sigma^S \sigma^C = \sigma^{S \bigtriangleup C}$
(\texttt{spinProduct\_mul}).
\end{proof}

\begin{lemma}[Iterated product --- \texttt{hasNonnegCorrelations\_finset\_prod},
\texttt{hasNonnegCorrelations\_mul\_prod}]
A product of factors $(a_k + b_k \sigma^{C_k})$ with $a_k, b_k \geq 0$
has HNC. More generally, multiplying an HNC function by such a product
preserves HNC.
\end{lemma}

\begin{proof}
By induction on the index set, applying Lemma~\ref{lem:mul} at each step.
\end{proof}

\subsection{Factoring the Boltzmann weight}

\begin{lemma}[\texttt{exp\_sign\_decomp}, \texttt{exp\_edgeSpin\_decomp}]
For $x \in \{+1, -1\}$ and $\alpha \geq 0$:
$e^{\alpha x} = \cosh\alpha + x\,\sinh\alpha$,
with $\cosh\alpha > 0$ and $\sinh\alpha \geq 0$.
\end{lemma}

\begin{lemma}[\texttt{bwFactored\_eq\_boltzmannWeight}]
\[
  e^{-\beta H_G(\sigma)}
  = \prod_{\{i,j\} \in E(G)} (\cosh(\beta J) + \sinh(\beta J)\,s_i s_j)
  \cdot \prod_{i \in \iota} (\cosh(\beta h) + \sinh(\beta h)\,s_i).
\]
\end{lemma}

\begin{proof}
$-\beta H = \beta J \sum_e s_i s_j + \beta h \sum_i s_i$.
Apply $\exp(\text{sum}) = \prod\exp$ (\texttt{Real.exp\_sum}),
then decompose each factor.
\end{proof}

\subsection{Conclusion}

\begin{theorem}[\texttt{boltzmannWeight\_hasNonnegCorrelations}]
$e^{-\beta H_G}$ has HNC for ferromagnetic parameters.
\end{theorem}

\begin{proof}
The factored form is a product of $(a+b\,\sigma^C)$ terms with
$a,b \geq 0$. Start from $f \equiv 1$ (HNC by base case),
multiply by each edge factor, then by each site factor. HNC is
preserved at each step by Lemma~\ref{lem:mul}.
\end{proof}

\noindent
GKS-I follows: $\langle \sigma^A \rangle
= Z^{-1}\sum_\sigma \sigma^A e^{-\beta H} \geq 0$
since $Z > 0$ and the numerator $\geq 0$ by HNC.

\section{GKS-II proof (\texttt{Inequalities/GKS.lean})}

The second Griffiths inequality is proved by the duplicate variable trick,
following Friedli--Velenik~\cite{friedli-velenik}, Theorem~3.49~(3.55),
pp.~127--128.

\begin{theorem}[GKS-II --- \texttt{gks\_second}]
For ferromagnetic parameters,
$\langle\sigma^A\sigma^B\rangle \geq \langle\sigma^A\rangle\langle\sigma^B\rangle$
for all $A, B$.
\end{theorem}

\subsection{Proof outline}

\begin{enumerate}
\item Introduce an independent copy $\chi$ of $\sigma$. Then
  \[
    Z^2\bigl(\langle\sigma^{A\triangle B}\rangle
      - \langle\sigma^A\rangle\langle\sigma^B\rangle\bigr)
    = \sum_{\omega,\omega'} \omega^A(\omega^B - \omega'^B)\,w(\omega)\,w(\omega').
  \]
  (\texttt{duplicateSum\_eq})

\item Change variables: $t_i = \omega_i \cdot \omega'_i$
  (\texttt{Spin.mul}). For fixed $\omega$, this is a bijection on
  $\mathrm{Config}\;\iota$. Under this substitution:
  \begin{itemize}
    \item $\omega^A(\omega^B - \omega'^B) = \omega^{A\triangle B}(1 - t^B)$
    \item $w(\omega)\,w(\omega\cdot t) = \mathrm{modifiedWeight}(t, \omega)$
      where each coupling $K_C$ is replaced by $K_C(1 + t^C) \geq 0$.
  \end{itemize}
  After swapping sums:
  \[
    \sum_t (1 - t^B) \sum_\omega \omega^{A\triangle B}\,
      \mathrm{modifiedWeight}(t, \omega).
  \]
  (\texttt{duplicateSum\_eq\_changed})

\item For each fixed $t$:
  \begin{itemize}
    \item $(1 - t^B) \geq 0$ since $t^B \in \{-1,+1\}$
      (\texttt{one\_sub\_spinProduct\_nonneg}).
    \item The inner sum $\geq 0$ by
      \texttt{hasNonnegCorrelations\_edge\_site\_product} applied to
      the modified couplings $K_C(1 + t^C)$
      (\texttt{modifiedWeight\_nonneg\_corr}).
  \end{itemize}

\item Since $Z^2 > 0$ and the duplicate sum $\geq 0$, GKS-II follows.
\end{enumerate}

\subsection{Key new definitions}

\begin{itemize}
  \item \texttt{Spin.mul}: spin multiplication ($\mathrm{up}$ acts as
    identity, $\mathrm{down}$ acts as flip).
    $\mathrm{toSign}(a \cdot b) = \mathrm{toSign}(a) \cdot \mathrm{toSign}(b)$.
  \item \texttt{modifiedWeight}: Boltzmann weight with couplings
    $K_C(1 + t^C)$.
  \item \texttt{hasNonnegCorrelations\_edge\_site\_product}: shared
    lemma for GKS-I and GKS-II (product of $\exp(K \cdot \text{spin})$
    factors with $K \geq 0$ has HNC).
\end{itemize}

\section{Numerical tests (\texttt{test/IsingModel/GKSTest.lean})}

A Float-based reference implementation verifies:
\begin{itemize}
  \item $Z > 0$ for 4 graph types
  \item Spin flip symmetry ($h = 0$)
  \item GKS-I for 6 parameter sets (2--4 sites, all subsets)
  \item GKS-II ($\langle\sigma^A\sigma^B\rangle \geq
    \langle\sigma^A\rangle\langle\sigma^B\rangle$)
    for 6 parameter sets (2--4 sites, all subset pairs)
  \item GKS-I and GKS-II violation for anti-ferromagnetic $J < 0$
\end{itemize}

\section{FKG inequality (\texttt{Inequalities/FKG.lean})}

The FKG inequality is proved by applying Mathlib's
\texttt{Combinatorics.SetFamily.FourFunctions.fkg} to the Ising Boltzmann
weight, following Friedli--Velenik~\cite{friedli-velenik}, Theorem~3.21
(p.~98) and Theorem~3.50 (pp.~128--130).

\begin{theorem}[\texttt{fkg\_ising}]
For ferromagnetic parameters and monotone nondecreasing nonneg
$f, g : \mathrm{Config}\;\iota \to \mathbb{R}$,
$\langle fg \rangle \geq \langle f \rangle \langle g \rangle$.
\end{theorem}

\subsection{Proof outline}
\begin{enumerate}
\item Define a \texttt{LinearOrder} on \texttt{Spin}: $\mathrm{down} < \mathrm{up}$
  ($-1 < +1$). The product order on $\mathrm{Config}\;\iota = \iota \to
  \mathrm{Spin}$ makes it a \texttt{DistribLattice}.
\item Prove log-supermodularity of the Boltzmann weight
  (\texttt{boltzmannWeight\_log\_supermodular}):
  $w(\sigma)w(\sigma') \leq w(\sigma \vee \sigma')w(\sigma \wedge \sigma')$.
  This reduces to the edge inequality
  $\sigma_i\sigma_j + \sigma'_i\sigma'_j \leq
  (\sigma_i \vee \sigma'_i)(\sigma_j \vee \sigma'_j) +
  (\sigma_i \wedge \sigma'_i)(\sigma_j \wedge \sigma'_j)$
  (\texttt{spin\_edge\_supermodular}, verified by exhaustive case analysis).
\item Apply Mathlib's \texttt{fkg} with $\mu = w$, then convert from
  unnormalized sums to \texttt{gibbsExpectation} form.
\end{enumerate}

\section{Asano Contraction}

The Asano contraction is the algebraic core of the Lee--Yang circle theorem.
It merges two variables in a multilinear polynomial, keeping only the
``both present'' and ``both absent'' parts
\cite[pp.~123--124]{friedli-velenik}.

\begin{definition}[Multilinear polynomial]
A multilinear polynomial over $\mathbb{C}$ with variables indexed by a
finite type $\iota$ is a function $p : \mathcal{P}(\iota) \to \mathbb{C}$,
evaluated as $p(z) = \sum_{X \subseteq \iota} p(X) \prod_{i \in X} z_i$.
In Lean: \texttt{Finset $\iota$ $\to$ $\mathbb{C}$}.
\end{definition}

\begin{definition}[Asano contraction]
Given $p$ and distinct $i, j \in \iota$, the contraction of $j$ into $i$
is $q(X) = p(X \cup \{j\})$ if $i \in X$ and $j \notin X$;
$q(X) = p(X)$ if $i \notin X$ and $j \notin X$;
$q(X) = 0$ if $j \in X$.
\end{definition}

\begin{theorem}[Asano contraction preserves non-vanishing]
\label{thm:asano}
If $p$ does not vanish on the open unit polydisk
$\{z : |z_k| < 1\;\forall k\}$, then neither does its Asano contraction.
Reference: \cite[Proposition~3.44]{friedli-velenik}.
\end{theorem}

The proof uses a bilinear non-vanishing lemma:

\begin{lemma}[Bilinear non-vanishing]
If $f(z,w) = azw + bw + cz + d \neq 0$ for all $|z|, |w| < 1$,
then $az + d \neq 0$ for all $|z| < 1$.
\end{lemma}

\begin{proof}[Proof sketch]
Show $\|a\| \leq \|d\|$ via the parallelogram law.
For $|z'| < 1$, the affine function $w \mapsto (az'+b)w + (cz'+d)$
has no zero with $|w| < 1$, giving $\|az'+b\| \leq \|cz'+d\|$.
Squaring and averaging at $z' = \pm t$ (real) via the parallelogram law:
$\mathrm{normSq}(a) \cdot t^2 + \mathrm{normSq}(b) \leq
\mathrm{normSq}(c) \cdot t^2 + \mathrm{normSq}(d)$.
Combining the $z$- and $w$-directions and applying a linear function argument
on $[0,1)$ yields $\mathrm{normSq}(a) \leq \mathrm{normSq}(d)$,
hence $\|a\| \leq \|d\|$. Then $az + d = 0$ would require
$|z| = |d/a| \geq 1$, contradicting $|z| < 1$.
\end{proof}

The contraction theorem follows by decomposing the polynomial evaluation
$p(z[i \to u, j \to w])$ as a bilinear form $auw + bw + cu + d$ in $(u, w)$,
where the contraction's evaluation equals $a \cdot z_i + d$.
The decomposition uses a product extraction lemma
(\texttt{prod\_upd}) and a sum reindexing via the involution
$X \mapsto X \triangle \{j\}$ (\texttt{Fintype.sum\_equiv}).

\begin{remark}
When Mathlib provides a native implementation of Asano contraction or
multilinear polynomials, the definitions here should be replaced.
\end{remark}

\section{Lee-Yang circle theorem (\texttt{LeeYang.lean})}

The Lee-Yang circle theorem states that the Ising partition polynomial
has no zeros on the open unit polydisk. Our formalization follows the
Harcos/Ruelle approach~\cite{harcos, ruelle}.

\begin{definition}[Lee-Yang polynomial]
For an $n\times n$ matrix $A$ over $\mathbb{C}$, define
\[
  f_A(z) = \sum_{S\subseteq [n]}
    \Bigl(\prod_{i\in S}\prod_{j\notin S} a_{ij}\Bigr)\prod_{k\in S} z_k.
\]
\end{definition}

\begin{theorem}[Ruelle~{\cite[Theorem~1, p.~589]{ruelle}};
  Harcos~{\cite[Theorem~3, p.~2]{harcos}};
  \texttt{leeYangPoly\_nonvanishing}]
If $A$ is Hermitian with $|a_{ij}|\le 1$, then $f_A(z)\ne 0$
for all $z$ with $|z_k|<1$.
\end{theorem}

The proof is by induction on $n$. The inductive step decomposes
$f_A(z)=\beta + z_n\cdot\alpha$ and shows $\|\alpha\|\le\|\beta\|$
via:
\begin{enumerate}
\item \textbf{Torus identity} (\texttt{torus\_identity}):
  On $|v_k|=1$,
  $\alpha(v) = (\prod v_k)\cdot\overline{\beta(v)}$,
  hence $\|\alpha\|=\|\beta\|$.
\item \textbf{Approximation}: Scale the last column by $t<1$
  so that $\beta_t\ne 0$ on the closed polydisk.
\item \textbf{Iterated max modulus} (\texttt{iterated\_ratio}):
  By single-variable maximum modulus
  (\texttt{one\_var\_max\_ratio}, via \texttt{DiffContOnCl}),
  iterate over each variable to get $\|\alpha_t/\beta_t\|\le 1$.
\item \textbf{Limit $t\to 1$}: Both sides are continuous in $t$;
  $\{t:\|\alpha_t\|\le\|\beta_t\|\}$ is closed and contains
  $[0,1)$, hence contains~$1$.
\end{enumerate}

The Ising application (\texttt{lee\_yang\_circle}) constructs a
Hermitian matrix from the edge list and coupling constants, reducing
the combinatorial Ising model to the matrix formulation.

\begin{thebibliography}{9}
\bibitem{glimm-jaffe}
J.~Glimm and A.~Jaffe,
\emph{Quantum Physics: A Functional Integral Point of View},
2nd ed., Springer, 1987.

\bibitem{ruelle}
D.~Ruelle,
\emph{Characterization of Lee-Yang polynomials},
Annals of Mathematics, 171(1):589--603, 2010.

\bibitem{harcos}
G.~Harcos,
\emph{The Lee-Yang Circle Theorem},
lecture notes based on Ruelle~\cite{ruelle},
\url{https://users.renyi.hu/~gharcos/lee-yang.pdf}.

\bibitem{friedli-velenik}
S.~Friedli and Y.~Velenik,
\emph{Statistical Mechanics of Lattice Systems:
A Concrete Mathematical Introduction},
Cambridge University Press, 2017.

\bibitem{simon}
B.~Simon,
\emph{The Statistical Mechanics of Lattice Gases, Vol.~I},
Princeton University Press, 1993.
\end{thebibliography}

\end{document}
